---
title: "Summary Data"
author: "Jeremy Mikecz"
format: html
editor: visual
jupyter: ir
---

# Get Summary Data (Practice with preloaded datasets)

## 1. Retrieve Pre-loaded Datasets

Base R and some of its core data science packages (like tidyverse and its dependencies) come with preloaded datasets. We can use these preloaded to practice our data science skills.

First, we must install and run **dplyr**:

```{r}
#install.packages("tidyverse")
library(dplyr)
```

to view the available preloaded datasets, run the following cell:

```{r}
#data()
```

In this lesson we will work with the **Star Wars** dataset (`starwars`) that loads with the **dplyr** library (part of the tidyverse package).

`{r} starwars}`

To learn more about this dataset run:

```{r}
help(starwars)
```

```{r}
starwars
```

## 2. Get Summary Data about a Dataframe

When working with an unfamiliar dataset, it helps to first extract some basic summary data from that dataset. Run the following commands and examine what each did.

```{r}
head(starwars)
#also try:
head(starwars, 3)
```

```{r}
dim(starwars) 
```

```{r}
colnames(starwars)
```

```{r}
table(starwars$species)
```

```{r}
xtabs(data=starwars, ~species + homeworld) |>
    as_tibble() |>
    filter(n > 0)
```

```{r}
#summary(starwars)

```

```{r}
str(starwars)
```

## 3. Retrieve Data

Get one row

```{r}
starwars[1,]
```

Get one cell

```{r}
starwars$name[4]
```

Get item in list

```{r}
print(starwars$starships[1]) 
starwars$starships[[1]][1]
```

::: tip
## Exercise

Choose another built-in dataset visible when you run `data()`. Use the commands above to explore the size, shape, type, and nature of the data stored in that dataset.
:::
